\section{Implementations}

This section details the method used to determine container sequences, as well as two heuristics that were implemented to fill the containers. By container sequence, we mean a list of the containers that will be used to pack items. For example, given container types 1, 2 and 3, a valid container sequence is $(1, 2, 2, 3, 3)$. Heuristics should try to put each item within the first container in the sequence. If that is not possible, they should try to place the item in the second container, and so on.

Determining a container sequence is a complex task: In the best scenario, the sequence is such that no items are left outside of a container, and the cost associated with the set of selected containers is minimal. Since it is computationally expensive to test every possible container sequence for a problem, we propose an integer programming (IP) model to find sequences that should at least come close to meeting these criteria. Afterwards, we adapt two heuristics from the literature to this method, which allows them to solve MCLPs.

\subsection{Container sequence determination via integer programming}

Suppose a set of $n$ cuboid container types of different widths ($W_k$), heights ($H_k$) and depths ($D_k$), with $k = 1,\dots,n$. We want to find a sequence of containers that all items can be packed into, with the possibility of repeating containers of each type, while also minimizing the costs associated with each container in the sequence. To do so, we propose the use of the IP presented in \cref{ip1,ip2,ip3,ip4}, where $c \in \mathbb{R}^n$ represents the cost of each container type, $s \in \mathbb{R}^n$ is the vector of decision variables that indicate how many units of each container to use, and $v \in \mathbb{R}^n$ is the vector containing the volume of each container ($v_i = W_iH_iD_i$).  

\begin{align}
        \min\ &c's& \label{ip1}\\
        \text{s.t. } &\sum_{i=1}^{n}v_is_i \geq V_s + V_c,& \label{ip2}\\
        &s_i \geq 0,\ &\forall i \in \{1,\dots,n\}, \label{ip3}\\
        &s_i \in \mathbb{Z},\ &\forall i \in \{1,\dots,n\}.
        \label{ip4}
\end{align}

\cref{ip1} minimizes the objective function, which is defined as the total cost of all selected containers. \cref{ip2} imposes that the total volume of the selected containers must be greater than the total volume of the items in stock ($V_s$) and the containers selected in a previous iteration of the IP model ($V_c$). At first, $V_c = 0$ and $V_s$ is the sum of all item volumes. Now suppose that the IP model is used to determine a sequence of containers, and the heuristic used is not able to load all the items into the selected containers. In this case, $V_c$ becomes the volume sum of all containers in the previous sequence, and $V_s$ becomes the volume sum of all items that the heuristic was unable to load. This ensures that the IP model searches for a new sequence of containers, since it must now determine a sequence of greater volume than before. Finally, \cref{ip3,ip4} establish that the containers are unitary.

With the values of $s$ determined, a sequence can be built by repeating container $i$ a total of $s_i$ times, for all $i \in \{1,\dots,n\}$. There are many ways to build this sequence, but we assume that it is better to start with the largest selected containers and finish with the smallest containers, since that might help to load larger items first, which can facilitate the packing process of smaller items later on.

\subsection{Wall-building}
\label{sec:wb implementation}

The wall-building (WB) heuristic attempts to load a single container by stacking cuboids within layers. A layer is a space with the same width and height as the container, but a lesser depth. To determine the depth of a layer, a box type is selected according to \cref{flow:primary box selection WBH}, where, for simplicity, \emph{box} is synonymous with \emph{box type}.

\input{tikz/wb_primary_box_selection.tex}

Once a box type is filtered, the feasible rotation with the greatest depth is selected, and the same depth is applied to the layer, as \cref{fig:WB layer example} shows. For unconstrained boxes, any of the six cuboid rotations is feasible. However, for boxes with a fixed height, only two rotations are feasible. \cref{fig:cuboid rotations} illustrates this.

\input{tikz/cuboid_rotations.tex}

From this point on, we refer to the box type that defines a layer's depth as its \emph{primary box type}. In its initial state, a layer contains a single empty cuboid space, which we call \emph{primary space}. Because they have the same depth, primary boxes are always loaded into primary spaces.

\input{tikz/wb_layer_depth_example.tex}

After the initial loading process, the primary space is divided into smaller cuboid spaces, which we call \emph{secondary spaces}. \cref{fig:remaining spaces first filling WBH A} illustrates this situation. In this example, two \emph{heightwise} spaces are created above the primary boxes, and one \emph{widthwise} space is created to their right. \cref{fig:remaining spaces first filling WBH B} shows the state of the layer after loading boxes into the widthwise space. In this case, one heightwise space is created on top of the secondary boxes, and one \emph{depthwise} space is created in front of them. The loading process is complete when none of the remaining spaces can be filled with the available box types. \cref{flow:secondary box selection WBH} is used to determine the box type for filling secondary spaces.

\input{tikz/wb_box_placements.tex}

\input{tikz/wb_secondary_box_selection.tex}

Before we discuss the loading procedure, we must introduce the concept of amalgamation. Suppose that after the layer in \cref{fig:remaining spaces first filling WBH} is fully loaded, the depthwise space created in \cref{fig:remaining spaces first filling WBH B} remains empty because no box type could fit in it. This depthwise space will be adjacent to the next layer in the container, which means it can be amalgamated with the next layer in an attempt to reduce wasted space. For instance, consider \cref{fig:amalgamation example}, which consists of the top view of two layers. The previous layer contains two empty depthwise spaces, which we assume to be at the same height or below the current layer (otherwise, no amalgamation is possible). Since space $\Omega$ has the least depth, any box we place within it is guaranteed not to overlap with other boxes from the previous layer, and thus we choose $\Omega$ to amalgamate with. After the amalgamation, the original space is split into two, $S_{L} = L$ and $S_{R} = R \cup \Omega$, with widths $w_L$ and $w_R$, respectively. However, changing these widths during loading might lead to better space utilization, which is why we introduce a \emph{flexible width} parameter, $\hat{w} = \phi w_{R}$, with $\phi \in [0,1]$. As we clarify further ahead (\cref{flow:box loading procedure WBH}), this parameter allows for the width of $S_L$ to grow to at most $w_L + \hat{w}$, with the width of $S_R$ decreasing accordingly.

\input{tikz/wb_amalgamation.tex}

We now describe the loading procedure. Let $B_{wh}$ be the box selected through \cref{flow:primary box selection WBH} or \cref{flow:secondary box selection WBH}, and $B_{hw}$ a rotation that swaps its width and height. If $B_{wh}$ has a fixed height, or if $B_{hw}$ does not fit in the space, then we keep $B_{wh}$. Otherwise, if enough stock exists to complete a column with either rotation, we select the rotation that results in the highest column. If neither rotation completes a column, we choose the rotation with the greatest height. This leads us to \cref{flow:box loading procedure WBH}, which describes how spaces are filled using the selected box type. Since spaces are filled by side-by-side columns of a single box rotation, determining the cuboid spaces that remain after loading is trivial.

\input{tikz/wb_loading_procedure.tex}

With this procedure, a single container can be loaded with items. To generalize the loading process to multiple containers, \cref{alg:wb mclp} is used.

\input{code/pseudo/wbmclp.tex}

\subsection{Genetic algorithm}

The GA developed is a simplified version of the one presented by \textcite{GONÇALVES2011}. To understand this method, we first need to enumerate the different item types from 1 to $n$. Then, let $q_1, \dots, q_n$ be the quantities of each item type. A non-decreasing sequence $S$ can be defined, containing item $k$ a total of $q_k$ times, for $k = 1,\dots,n$. For example, given 3 item types with $q_1 = q_2 = 2$, $q_3 = 1$, it follows that $S = (1, 1, 2, 2, 3)$.

For a problem with $Q = \sum_{k=1}^{n}q_k$ items, a population of chromosomes with $2Q$ genes is generated, with values in the $[0, 1]$ interval. The first $Q$ genes specify the order in which the heuristic attempts to place items. This is done by associating each gene to the item type with the same index in $S$, and then reordering $S$ the same way as needed to put the first $Q$ genes in ascending order. As an example, suppose $S = (1, 1, 2, 2, 3)$, and a chromosome whose first five genes are $(0.93, 0.42, 0.17, 0.48, 0.80)$. If we rearrange these numbers in ascending order, we obtain $(0.17, 0.42, 0.48, 0.80, 0.93)$. By equivalently swapping the items in $S$, we get its rearrangement, $\bar{S} = (2, 1, 2, 3, 1)$.

The remaining $Q$ genes inform how each item must be placed. Specifically, gene $g_{Q+k}$ is used to determine the rotation and plane of item $\bar{S}_k$, $k = 1, \dots, Q$. Valid rotations for an item can be either variable- or fixed-height rotations, as shown in \cref{fig:cuboid rotations}. The plane can be $xy$, $xz$ or $yz$, and it indicates the axes along which the item is placed. \cref{fig:xy plane example} illustrates the filling of a cuboid space along the $xy$ plane, supposing that there are four items to place. The process consists of selecting one of the axes, $x$ or $y$, and attempting to fill it with as many strips of items as possible, such that the resulting packing is a cuboid. In the first packing, the $x$ plane is prioritized. As a result, one horizontal strip with two items is placed. Since there is enough vertical and horizontal space left, another strip of two boxes is placed on top of the first. In the second packing, the $y$ plane is prioritized. Three items can be stacked vertically, which leaves one. This remaining item is not enough to complete another column, therefore it is not placed.

\input{tikz/xy_plane_example.tex}

Since there are at most six rotations, and each plane can be filled in two different ways, there is a maximum of 36 possible ways to place an item in a given space. In order to select one of the item configurations, all possibilities are mapped to different subintervals of equal length of $[0,1]$. Then, whichever interval the respective gene belongs to is used to define the item's placement.

To select the space in which to place an item, the back-bottom-left procedure (BBL) is used. Let $E_k$ be the $k$-th empty space available for packing. BBL orders spaces such that $E_i < E_j$ if $x_i < x_j$, or $x_i = x_j$ and $y_i < y_j$, or $x_i = x_j$ and $y_i = y_j$ and $z_i < z_j$. Then, the first space in the ordering where the item fits is selected. After an item is placed, new empty cuboid spaces are created, adjacent to the item. To calculate these spaces, the method described by \textcite{LAI1997} is used, with one difference: The space on top of the item is limited to the area the item occupies. This is to guarantee the vertical stability of items, by ensuring that they are either placed on the ground or that their base area is completely supported by an item directly below.

In order to extend the GA by \textcite{GONÇALVES2011} to multiple containers, the BBL procedure is applied to the first container in the sequence. If the current item does not fit in any of the remaining spaces, the next container in the sequence is selected, and the BBL procedure is applied again. This process is repeated until a fitting space is found, or there are no containers left to check.

A chromosome's fitness is calculated as follows: Let $V_{Ti}$ and $V_{Fi}$ be, respectively, the total and the filled volume of container $i$. Let $k$ be the number of containers in the sequence determined by the IP model. We first calculate the fitness of a given chromosome as:

\begin{equation}
    F = \frac{100}{k}\sum_{i=1}^{k}\frac{V_{Fi}}{V_{Ti}}.
\end{equation}

Which gives us the mean percentage of filled volume per container. Now let $L$ be the number of items left out of the containers after the packing process is finished. If $L > 0$, a penalty must be applied to the fitness value. In our case, we chose

\begin{equation}
    \bar{F} = \frac{F}{10L}
\end{equation}
as the new fitness value, because it greatly decreases with the number of items that are left outside of the containers.

Once a population is generated and the fitness value of each solution is determined, the chromosomes are sorted according to their fitness values. A certain subset of the population, containing the best fitness values, is immediately copied to the next population, so as to ensure that the best fitness does not decrease from each iteration to the next. This subset of the population is called its \emph{elite}.

Another part of the next population is made of crossovers between solutions from the elite and solutions from the whole population, both chosen at random. Uniform crossover is used to generate a single offspring, with a higher chance of selecting genes from the elite chromosome. Finally, the last part of the next population is created by randomly generating as many chromosomes as needed to match the number of solutions in the current population. This can be seen as an extreme case of mutation.